# Reanalysis of Scriver et al.
This notebook contains a reanalysis of free-eDNA measurements from the study "Drop it all: extraction-free detection of targeted marine species through optimized direct droplet digital PCR" by Scriver et al. (https://doi.org/10.7717/peerj.16969).

## Imports
```{r}
library(dplyr)
library(tidyr)
library(dPCRfit)
library(ggplot2)
library(ggdist)
library(data.table)
```

```{r}
source("~/Documents/dev/wastewater-generative/code/utils_dists.R")
source("~/Documents/dev/wastewater-generative/code/utils_dPCR_statistics.R")
```

## Data
```{r}
aquarium <- readxl::read_excel(here::here("data","eDNA","aquarium_ddPCR_raw_data.xlsx"), sheet = 1)
```

```{r}
str(aquarium)
```

```{r}
conc_data_all <- aquarium |> 
  filter(Organism_Present == "No", Tank != "neg", Target_Species == "Bugula_neritina") |>
  mutate(
    Biomass = factor(Biomass, levels = c("high", "medium", "low"), ordered = TRUE),
    sample_position = stringr::str_extract(
      pattern = "(?<=\\d_.\\d.)[0-9]+", Sample_ID # add sample position (based on sample ID)
      ), 
    n_averaged = 1,
    )
```

Add partition information
```{r}
dPCR_partitions <- sapply(conc_data_all$Concentration, backcalc_partitions, c = 0.85*0.001, partitions_mean = 20000)
conc_data_all$total_partitions <- as.numeric(dPCR_partitions["total",])
conc_data_all$positive_partitions <- as.numeric(dPCR_partitions["positive",])
conc_data_all$total_partitions[is.na(conc_data_all$total_partitions)] <- rnbinom2(
  n = sum(is.na(conc_data_all$total_partitions)),
  mean = mean(conc_data_all$total_partitions, na.rm = T), sd = sd(conc_data_all$total_partitions, na.rm = T)
  )
```

## Regression

```{r}
rerun_models <- FALSE
model_types <- c("dPCR_binomial", "Averaged", "Normal", "Log-Normal", "dPCR")
biomass_levels <- c("all", "low", "medium", "high")
```

```{r}
fit_opts <- set_fit_opts(
  sampler = sampler_stan_mcmc(
    seed = 0, chains = 4, parallel_chains = 4, iter_warmup = 1000, iter_sampling = 1000
    )
  )
```

```{r}
if (rerun_models) {
  all_fits <- mapply(function(model_type, biomass_level) {
    
    if (biomass_level == "all") {
      setDT(conc_data_all)
      conc_data <- data.table::copy(conc_data_all)
      conc_data <- conc_data[, Biomass := factor(Biomass, ordered = FALSE)]
      reg_formula <- Concentration ~ Hours + Biomass + sample_position
    } else {
      conc_data <- conc_data_all |> filter(Biomass == biomass_level)
      reg_formula <- Concentration ~ Hours + sample_position
    }
    
    if (model_type == "Averaged") {
      conc_data <- conc_data |> 
          group_by(Target_Species, Hours) |> 
          filter(Concentration > 0.08) |> 
          summarize(Concentration = mean(Concentration), n_averaged = sum(n_averaged)) |> 
          mutate(Sample_ID = paste0("s_", Hours))
      reg_formula <- Concentration ~ Hours
    }
    
    if (model_type %in% c("Averaged", "Normal")) {
      fit <- dPCRfit(
        formula = reg_formula,
        data = conc_data,
        link = "log",
        prior_intercept = c(0, 0.5),
        prior_coefficients = c(0, 0.5),
        measurements = concentration_measurements(
          id_col = "Sample_ID",
          concentration_col = "Concentration",
          n_averaged_col = "n_averaged",
          distribution = "normal"
        ),
        noise = noise_constant_var(
          cv_prior_mu = 0, cv_prior_sigma = 0.5, warn = FALSE
        ),
        nondetect = nondetect_none(),
        fit_opts = fit_opts
      )
    } else if (model_type == "Log-Normal") {
      fit <- dPCRfit(
        formula = reg_formula,
        data = conc_data,
        link = "log",
        prior_intercept = c(0, 0.5),
        prior_coefficients = c(0, 0.5),
        measurements = concentration_measurements(
          id_col = "Sample_ID",
          concentration_col = "Concentration",
          n_averaged_col = "n_averaged",
          distribution = "lognormal"
        ),
        noise = noise_constant_cv(
          cv_prior_mu = 0, cv_prior_sigma = 0.5
        ),
        nondetect = nondetect_none(),
        fit_opts = fit_opts
      )
    } else if (model_type == "dPCR") {
      fit <- dPCRfit(
        formula = reg_formula,
        data = conc_data,
        link = "log",
        prior_intercept = c(0, 0.5),
        prior_coefficients = c(0, 0.5),
        measurements = concentration_measurements(
          id_col = "Sample_ID",
          concentration_col = "Concentration",
          n_averaged_col = "n_averaged"
        ),
        noise = noise_dPCR(
          cv_prior_mu = 0, cv_prior_sigma = 0.5,
          max_partitions_prior_lower = 10000,
          max_partitions_prior_upper = 50000,
          partition_loss_mean_prior_lower = 0.01,
          partition_loss_mean_prior_upper = 0.3,
          partition_loss_variation_prior_lower = 0.5,
          partition_loss_variation_prior_upper = 2,
          partition_loss_max = 0.5,
          volume_scaled_prior_mu = 1e-3,
          volume_scaled_prior_sigma = 0.5e-3,
          prePCR_noise_type = "gamma"
        ),
        nondetect = nondetect_dPCR(),
        fit_opts = fit_opts
      )
    } else if (model_type == "dPCR_binomial") {
      fit <- dPCRfit(
        formula = reg_formula,
        data = conc_data,
        link = "log",
        prior_intercept = c(0, 0.5),
        prior_coefficients = c(0, 0.5),
        measurements = positive_partitions(
          id_col = "Sample_ID",
          positive_partitions_col = "positive_partitions",
          total_partitions_col = "total_partitions",
          n_averaged_col = "n_averaged"
        ),
        noise = noise_dPCR(
          cv_prior_mu = 0, cv_prior_sigma = 0.5,
          partitions_observe = TRUE,
          volume_scaled_prior_mu = 0.85e-3,
          volume_scaled_prior_sigma = 0,
          prePCR_noise_type = "gamma"
        ),
        nondetect = nondetect_dPCR(),
        fit_opts = fit_opts
      )
    }
    return(list(fit = fit, model_type = model_type, biomass_level = biomass_level))
  },
  model_type = rep(model_types, length(biomass_levels)),
  biomass_level = rep(biomass_levels, each = length(model_types)),
  SIMPLIFY = FALSE
  )
}
```

## Diagnostics
```{r}
if (rerun_models) {
  diags <- suppressMessages(rbindlist(lapply(all_fits, function(res) {
    diags <- res$fit$fit$diagnostic_summary()
    iters <- res$fit$fit_opts$sampler$iter_sampling + res$fit$fit_opts$sampler$iter_warmup
    data.frame(
      model_type = factor(res$model_type, levels = c("Averaged", "Normal", "Log-Normal", "dPCR", "dPCR_binomial")),
      biomass_level = factor(res$biomass_level, levels = c("low", "medium", "high", "all"), ordered = TRUE),
      percent_divergent = sum(diags$num_divergent)/iters*100,
      percent_max_treedepth = sum(diags$num_max_treedepth)/iters*100,
      min_ebfmi = min(diags$ebfmi)
      )
  })))
  saveRDS(diags, here::here("data", "results", "aquarium_Scriver_et_al_diagnostics.rds"))
} else {
  diags <- readRDS(here::here("data", "results", "aquarium_Scriver_et_al_diagnostics.rds"))
}
print(diags)
```

## Results
```{r}
if (rerun_models) {
  regression_summary <- rbindlist(lapply(all_fits, function(res) {
    coef_summary <- res$fit$coef_summary
    coef_summary[, model_type := factor(res$model_type, levels = c("Averaged", "Normal", "Log-Normal", "dPCR", "dPCR_binomial"))]
    coef_summary[, biomass_level := factor(res$biomass_level, levels = c("low", "medium", "high", "all"), ordered = TRUE)]
    setcolorder(coef_summary, c("model_type", "biomass_level", "variable"))
  }))
  saveRDS(regression_summary, here::here("data", "results", "aquarium_Scriver_et_al_regression_summary.rds"))
  
  regression_half_life <- bind_rows(lapply(all_fits, function(res) {
    tidybayes::spread_draws(res$fit$fit, beta[i]) |> 
      filter(i == 1) |> 
      mutate(half_life = -log(2)/beta) |> 
      ungroup() |>
      summarize(
        mean = mean(half_life),
        median = median(half_life),
        sd = sd(half_life),
        lower_95 = quantile(half_life, p = 0.025),
        upper_95 = quantile(half_life, p = 0.975)
      ) |> 
      mutate(
        model_type = factor(res$model_type, levels = c("Averaged", "Normal", "Log-Normal", "dPCR", "dPCR_binomial")),
        biomass_level = factor(res$biomass_level, levels = c("low", "medium", "high", "all"), ordered = TRUE),
        .before = 1)
  }))
  saveRDS(regression_half_life, here::here("data", "results", "aquarium_Scriver_et_al_regression_half_life.rds"))
  
  regression_linear <- rbindlist(lapply(all_fits, function(res) {
    if (res$model_type == "Averaged") {
      newdata <- data.frame(
        Hours = min(conc_data_all$Hours):max(conc_data_all$Hours)
        )
    } else {
        if (res$biomass_level == "all") {
        newdata <- data.frame(
          Hours = min(conc_data_all$Hours):max(conc_data_all$Hours),
          Biomass = "high",
          sample_position = 1
          )
      } else {
        newdata <- data.frame(
          Hours = min(conc_data_all$Hours):max(conc_data_all$Hours),
          sample_position = 1
          )
      }
    }
    
    preds <- predict(
      res$fit,
      newdata = newdata,
      interval = "confidence",
      keep_data = TRUE
      )
    preds[, model_type := factor(res$model_type, levels = c("Averaged", "Normal", "Log-Normal", "dPCR", "dPCR_binomial"))]
    preds[, biomass_level := factor(res$biomass_level, levels = c("low", "medium", "high", "all"), ordered = TRUE)]
    setcolorder(preds, c("model_type", "biomass_level"))
  }), fill = TRUE)
  saveRDS(regression_linear, here::here("data", "results", "aquarium_Scriver_et_al_regression_linear.rds"))
} else {
  regression_summary <- readRDS(here::here("data", "results", "aquarium_Scriver_et_al_regression_summary.rds"))
  regression_half_life <- readRDS(here::here("data", "results", "aquarium_Scriver_et_al_regression_half_life.rds"))
  regression_linear <- readRDS(here::here("data", "results", "aquarium_Scriver_et_al_regression_linear.rds"))
}
```

### Coefficients

```{r}
regression_summary |> 
  filter(variable %in% c("(Intercept)","Hours"))
```

### Half-life estimates

```{r, fig.width=8, fig.height=3}
halflife_plot <- regression_half_life |> 
  mutate(partitions_known = ifelse(model_type == "dPCR_binomial", "Partitions\n available", "Partitions\n not available")) |> 
  filter(model_type != "Averaged", biomass_level != "all") |>
  mutate(biomass_level = factor(biomass_level, levels = c("low", "medium", "high"), labels = rev(c("High biomass", "Medium biomass", "Low biomass")), ordered = TRUE)) |> 
  mutate(model_type = factor(
    model_type,
    levels = rev(c("dPCR_binomial", "dPCR", "Normal", "Log-Normal")),
    labels = rev(c("Binomial", "dPCR", "Normal", "Log-Normal")),
    ordered = T)
    ) |> 
  ggplot(aes(y = model_type, x = mean, color = model_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 16.86489, linetype = "dotted", color = "darkred") +
  geom_pointinterval(aes(xmin = lower_95, xmax = upper_95, x = median), size = 8, linewidth = 2.5) +
  scale_x_continuous(breaks = seq(-200,200,20)) +
  coord_cartesian(xlim = c(-2, 65)) +
  facet_grid(partitions_known~biomass_level, scales = "free_y", ) +
  #force_panelsizes(rows = c(1, 1.5)) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.title.y = element_blank(),
    strip.background = element_rect(fill = "#f2f2f2")
    ) +
  xlab("Half-life [hours]") +
  scale_color_manual(values = rev(c("darkgrey","#007191","#d31f11","#f47a00")))
halflife_plot
```

```{r}
regression_half_life |> filter(model_type == "dPCR_binomial") |> select(model_type, biomass_level, median, lower_95, upper_95)
regression_half_life |> filter(model_type == "dPCR") |> select(model_type, biomass_level, median, lower_95, upper_95)
regression_half_life |> filter(model_type == "Log-Normal") |> select(model_type, biomass_level, median, lower_95, upper_95)
regression_half_life |> filter(model_type == "Normal") |> select(model_type, biomass_level, median, lower_95, upper_95)
```

### Model fit
```{r}
measurements_to_plot <- conc_data_all |> rename(biomass_level = Biomass)
regression_linear_to_plot <- regression_linear[model_type != "Averaged",]
regression_linear_to_plot <- regression_linear_to_plot[biomass_level != "all",]
regression_linear_to_plot[, model_type := factor(model_type, levels = c("dPCR_binomial", "dPCR", "Normal", "Log-Normal"), labels = c("Binomial", "dPCR", "Normal", "Log-Normal"), ordered = T)]

model_color_palette <- c("Binomial" = "darkgrey", "dPCR" = "#007191", "Normal" = "#d31f11", "Log-Normal" = "#f47a00")

conc_plot <- ggplot(measurements_to_plot) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_ribbon(data = regression_linear_to_plot, aes(x = Hours, ymin = lower_95, ymax = upper_95, fill = model_type), alpha = 0.2) +
  geom_line(data = regression_linear_to_plot, aes(x=Hours, y = mean, color = model_type)) +
  geom_jitter(aes(x=Hours, y=Concentration), color = "black", size = 0.75, width = 0) +
  facet_wrap(~factor(biomass_level, levels = rev(c("high", "medium", "low")), labels = rev(c("High biomass", "Medium biomass", "Low biomass"))), scales = "free_y") +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    strip.background = element_blank(),
    strip.text = element_blank(),
    ) +
  xlab("Time since removal of organism [hours]") +
  ylab(expression(paste("Concentration [gc/", mu, "L]"))) +
  coord_cartesian(xlim = c(-2, 70)) +
  scale_fill_manual(values = model_color_palette) +
  scale_color_manual(values = model_color_palette)
  #scale_y_log10()
conc_plot
```

### Combined figure
```{r}
cowplot::plot_grid(
  halflife_plot,
  conc_plot + theme(axis.title.y = element_blank(), plot.background = element_blank(), legend.position = "None"), ncol = 1, align = "v", axis = "lrtb")
```


```{r, fig.width = 10, fig.height = 6}
main_grid <- cowplot::plot_grid(
  cowplot::plot_grid(NULL, cowplot::get_plot_component(conc_plot, pattern = "guide-box-top"), nrow = 1, rel_widths = c(0.07, 0.93)),
  cowplot::plot_grid(
  halflife_plot + theme(plot.background = element_blank()),
  conc_plot + theme(axis.title.y = element_blank(), plot.background = element_blank(), legend.position = "None"),
  ncol = 1, rel_heights = c(0.45, 0.55), align = "v", axis = "lr", labels = "AUTO"),
  ncol = 1, rel_heights = c(0.07, 0.93), align = "v", axis = "l"
)

cowplot::plot_grid(
  NULL,
  cowplot::plot_grid(NULL, cowplot::get_plot_component(conc_plot, "ylab-l"), ncol = 1, rel_heights = c(0.45, 0.55)), 
  NULL,
  main_grid,
  nrow = 1, rel_widths = c(0.025, 0.02, -0.05, 0.98)
  )

ggsave(here::here("~/Documents/dev/wastewater-generative", "figures", "dPCR paper", "aquarium_Bugula_neritina.pdf"), width = 10, height = 6)
```
