---
title: "Validation of theoretical predictions for CV and probability of non-detection via simulation"
---

## Preparation
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
here::i_am("notebooks/dPCR paper/CV_p_nondetect_sim.Rmd")
```

```{r}
source(here::here("code", "utils_dPCR_statistics.R"))
source(here::here("code", "utils_dists.R"))
```

```{r}
rerun_computations <- FALSE
```


## Coefficient of variation

### Helper functions

Computation of coefficient of variation (via simulation and analytical approximations)
```{r}
compute_dat_cv_pre <- function(example_n_replicates, example_c, example_ndroplets, example_cv_values, example_lambda_values, simulate = FALSE, seed = 0) {
  set.seed(seed)
  
  dat_cv_pre <- expand.grid(example_lambda_values, example_c, example_ndroplets, example_cv_values, example_n_replicates)
  names(dat_cv_pre) <- c("lambda", "c", "ndroplets", "cv", "n_replicates")

  # simulation
  if (simulate) {
    # simulation for gamma
    dat_cv_pre$conc_cv_sim_gamma = parallel::mcmapply(sim_cv_pre,
      la = dat_cv_pre$lambda, c = dat_cv_pre$c, n = dat_cv_pre$ndroplets,
      cv = dat_cv_pre$cv, n_replicates = dat_cv_pre$n_replicates,
      distribution = "gamma", n_samples = 1e04, mc.cores = 8
      )
  
    # simulation for lognormal
  dat_cv_pre$conc_cv_sim_lnorm = parallel::mcmapply(sim_cv_pre,
    la = dat_cv_pre$lambda, c = dat_cv_pre$c, n = dat_cv_pre$ndroplets,
    cv = dat_cv_pre$cv, n_replicates = dat_cv_pre$n_replicates,
    distribution = "lnorm", n_samples = 1e04, mc.cores = 8
    )
  
  }

  # gamma
  dat_cv_pre$conc_cv_gamma = mapply(est_cv_pre_gamma,
    la = dat_cv_pre$lambda, c = dat_cv_pre$c, n = dat_cv_pre$ndroplets,
    cv = dat_cv_pre$cv, n_replicates = dat_cv_pre$n_replicates
    )
  
  # # log-normal
  dat_cv_pre$conc_cv_lnorm = mapply(est_cv_pre_lnorm,
    la = dat_cv_pre$lambda, c = dat_cv_pre$c, n = dat_cv_pre$ndroplets,
    cv = dat_cv_pre$cv, n_replicates = dat_cv_pre$n_replicates
    )
  
  # taylor series approximation
  dat_cv_pre$conc_cv_taylor = mapply(est_cv_pre_taylor,
    la = dat_cv_pre$lambda, c = dat_cv_pre$c, n = dat_cv_pre$ndroplets,
    cv = dat_cv_pre$cv, n_replicates = dat_cv_pre$n_replicates
    )
  
  dat_cv_pre$cv = as.factor(dat_cv_pre$cv)

  return(dat_cv_pre)
}
```

```{r}
format_cv_plot <- function(p) {
  p +
    scale_x_continuous(expand = c(0,0)) +
    facet_wrap(~n_replicates_label) + 
    theme_bw() +
    xlab("Concentration in sample [gc/mL]") + 
    ylab("Coefficient of variation") +
    scale_color_discrete(name = "Pre-PCR coefficient of variation") + 
    scale_y_log10() +
    theme(
     legend.position = "top",
     strip.background = element_rect(fill = "#f2f2f2")
    ) +
    coord_cartesian(ylim = c(NA, 7))
}
```

### Main example
```{r}
# simulate/compute CV, this can take a while
if (rerun_computations) {
  dat_cv_pre <- compute_dat_cv_pre(
    example_n_replicates = c(1,2),
    example_c = 0.519e-6 * 30,
    example_ndroplets = 25000,
    example_cv_values = c(0, 0.5, 1, 1.5, 2),
    example_lambda_values = seq(0.001, 15, 0.1),
    simulate = TRUE,
    seed = 0
  )
  dat_cv_pre <- dat_cv_pre |> 
    mutate(
      n_replicates_label = paste(
        ifelse(n_replicates==1,"(i)","(ii)"), n_replicates, ifelse(
          n_replicates==1,"replicate","replicates")
        )
      )
  saveRDS(dat_cv_pre, here::here("data", "results", "dat_cv_pre.rds"))
}
```

```{r}
dat_cv_pre <- readRDS(here::here("data", "results", "dat_cv_pre.rds"))
```

#### Log-normal noise, Taylor approximation
```{r} 
(ggplot(dat_cv_pre, aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_lnorm, color = cv, group = cv), alpha = 0.5) + 
  geom_line(aes(y  = conc_cv_taylor, color = cv, group = cv))) |> 
  format_cv_plot()

ggsave(here::here("figures", "dPCR paper", "cv_pre_lnorm_taylor.pdf"), width = 10, height = 5)
```

#### Log-normal noise, MGF approximation using Lambert function
```{r} 
(ggplot(dat_cv_pre, aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_lnorm, color = cv, group = cv), alpha = 0.5) + 
  geom_line(aes(y  = conc_cv_lnorm, color = cv, group = cv))) |> 
  format_cv_plot()

ggsave(here::here("figures", "dPCR paper", "cv_pre_lnorm_lambert.pdf"), width = 10, height = 5)
```

#### Log-normal noise, comparison of MGF and Taylor approximations

```{r} 
(ggplot(dat_cv_pre, aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_lnorm, color = cv, group = cv), alpha = 0.5) + 
   geom_line(aes(y = conc_cv_taylor, color = cv, group = cv), linetype = "solid") +  geom_line(aes(y  = conc_cv_lnorm, color = cv, group = cv), linetype = "dashed")
 ) |> 
  format_cv_plot()

ggsave(here::here("figures", "dPCR paper", "cv_pre_lnorm_comparison.pdf"), width = 10, height = 5)
```

#### Gamma noise, MGF solution
```{r} 
cv_pre_plot_gamma <- (ggplot(dat_cv_pre, aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_gamma, color = cv, group = cv), alpha = 0.5) + 
  geom_line(aes(y  = conc_cv_gamma, color = cv, group = cv))) |> 
  format_cv_plot()

ggsave(cv_pre_plot_gamma, filename = here::here("figures", "dPCR paper", "cv_pre_gamma.pdf"), width = 10, height = 5)
```

#### Gamma noise, Taylor approximation
```{r} 
(ggplot(dat_cv_pre, aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_gamma, color = cv, group = cv), alpha = 0.5) + 
  geom_line(aes(y  = conc_cv_taylor, color = cv, group = cv))) |> 
  format_cv_plot()

ggsave(here::here("figures", "dPCR paper", "cv_pre_gamma_taylor.pdf"), width = 10, height = 5)
```

### Sensitivity analyses

#### Extreme range
```{r}
if (rerun_computations) {
  dat_cv_pre_wide <- compute_dat_cv_pre(
    example_n_replicates = c(1,2),
    example_c = 0.519e-6 * 30,
    example_ndroplets = 25000,
    example_cv_values = c(0, 0.05, 0.1, 0.15),
    example_lambda_values <- c(
      seq(0.001, 2.9, 0.1),
      seq(3, 99, 1),
      seq(100, 9000, 100),
      seq(9000, 900000, 1000)
      ),
    simulate = FALSE, # here we only compare theoretical predictions
    seed = 0
  )
  
  dat_cv_pre_wide <- dat_cv_pre_wide |> 
    mutate(
      n_replicates_label = paste(
        ifelse(n_replicates==1,"(i)","(ii)"), n_replicates, ifelse(
          n_replicates==1,"replicate","replicates")
        )
      )
  
  saveRDS(dat_cv_pre_wide, here::here("data", "results", "dat_cv_pre_wide.rds"))
}
```

```{r}
dat_cv_pre_wide <- readRDS(here::here("data", "results", "dat_cv_pre_wide.rds"))
```

```{r} 
(ggplot(dat_cv_pre_wide, aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_lnorm, color = cv, group = cv), linetype = "dotted") +
  geom_line(aes(y = conc_cv_taylor, color = cv, group = cv), linetype = "dashed") +
  geom_line(aes(y  = conc_cv_gamma, color = cv, group = cv))
  ) |> 
  format_cv_plot()

ggsave(here::here("figures", "dPCR paper", "cv_pre_wide.pdf"), width = 10, height = 5)
```

#### Conversion factor c
```{r}
if (rerun_computations) {
  dat_cv_pre_sensitivity_c <- compute_dat_cv_pre(
    example_n_replicates = c(1,2),
    example_c = c(0.519e-6 * 30, 0.519e-6 * 30 / 10, 0.519e-6 * 30 * 10),
    example_ndroplets = 25000,
    example_cv_values = c(0.5),
    example_lambda_values = seq(0.001, 15, 0.1),
    simulate = TRUE,
    seed = 0
  )
  
  dat_cv_pre_sensitivity_c <- dat_cv_pre_sensitivity_c |> 
      mutate(
        n_replicates_label = paste(
          ifelse(n_replicates==1,"(i)","(ii)"), n_replicates, ifelse(
            n_replicates==1,"replicate","replicates")
          )
        )
  
  saveRDS(dat_cv_pre_sensitivity_c, here::here("data", "results", "dat_cv_pre_sensitivity_c.rds"))
}
```

```{r}
dat_cv_pre_sensitivity_c <- readRDS(here::here("data", "results", "dat_cv_pre_sensitivity_c.rds"))
```

```{r}
(dat_cv_pre_sensitivity_c |> 
  mutate(,
    c = factor(c, levels = c(1.557e-05, 1.557e-04, 1.557e-06), labels = c("1.557e-05", "1.557e-05 x10", "1.557e-05 x0.1"))
    ) |> 
ggplot(aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_gamma, color = c, group = c), alpha = 0.5) + 
  geom_line(aes(y  = conc_cv_gamma, color = c, group = c))) |> 
  format_cv_plot() +
  scale_color_discrete(name = "Conversion factor")

ggsave(here::here("figures", "dPCR paper", "cv_pre_sensitivity_c.pdf"), width = 10, height = 5)
```

#### Total number of partitions
```{r}
if (rerun_computations) {
  dat_cv_pre_sensitivity_m <- compute_dat_cv_pre(
    example_n_replicates = c(1,2),
    example_c = 0.519e-6 * 30,
    example_ndroplets = c(25000, 10000, 40000),
    example_cv_values = c(0.5),
    example_lambda_values = seq(0.001, 15, 0.1),
    simulate = TRUE,
    seed = 0
  )
  
  dat_cv_pre_sensitivity_m <- dat_cv_pre_sensitivity_m |> 
      mutate(
        n_replicates_label = paste(
          ifelse(n_replicates==1,"(i)","(ii)"), n_replicates, ifelse(
            n_replicates==1,"replicate","replicates")
          )
        )
  
  saveRDS(dat_cv_pre_sensitivity_m, here::here("data", "results", "dat_cv_pre_sensitivity_m.rds"))
}
```

```{r}
dat_cv_pre_sensitivity_m <- readRDS(here::here("data", "results", "dat_cv_pre_sensitivity_m.rds"))
```

```{r}
(dat_cv_pre_sensitivity_m |> 
  mutate(
    ndroplets = factor(ndroplets, levels = c(25000, 10000, 40000))
    ) |> 
ggplot(aes(x=lambda)) + 
  geom_line(aes(y = conc_cv_sim_gamma, color = ndroplets, group = ndroplets), alpha = 0.5) + 
  geom_line(aes(y  = conc_cv_gamma, color = ndroplets, group = ndroplets))) |> 
  format_cv_plot() +
  scale_color_discrete(name = "Total number of droplets")

ggsave(here::here("figures", "dPCR paper", "cv_pre_sensitivity_m.pdf"), width = 10, height = 5)
```

## Probability of non-detection

### Helper functions

Computation of probability of non-detection (via simulation and analytical approximations)
```{r}
compute_nondetection_probs <- function(example_lambda_values, example_c, example_total_partitions, example_cv, noise_dist, example_replicates, seed = 0) {
set.seed(seed)
nondetection_probs <- bind_rows(lapply(example_replicates, function(n_replicates) {
  nondetection_probs <- bind_rows(lapply(example_cv, function(technical_cv) {
    nondetection_probs <- data.frame(cv = technical_cv, lambda = example_lambda_values)
    nondetection_probs$simulated <- parallel::mcmapply(sim_nondetection_prob_pre,
      lambda = nondetection_probs$lambda,
      c = example_c,
      n=example_total_partitions,
      cv = technical_cv,
      n_replicates = n_replicates,
      noise_dist = noise_dist,
      mc.cores = 4
      )
    nondetection_probs$theoretical <- sapply(nondetection_probs$lambda, est_nondetection_prob_pre,
      c = example_c, n=example_total_partitions,
      cv = technical_cv,
      n_replicates = n_replicates,
      noise_dist = noise_dist
      )
    return(nondetection_probs)
  }))
  nondetection_probs$cv <- as.factor(nondetection_probs$cv)
  nondetection_probs$n_replicates <- n_replicates
  return(nondetection_probs)
}))
return(nondetection_probs)
}
```

```{r}
format_nondetect_plot <- function(p) {
  p +
    geom_hline(yintercept = 0.05, color = "black", linetype = "dashed") +
    xlab("Concentration in sample [gc/mL]") + ylab("Non-detection probability") + 
    theme_bw() + 
    facet_wrap(~n_replicates_label) + 
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(labels = scales::percent, expand = expansion(add = c(0, 0))) +
     coord_cartesian(ylim = c(0, 1), clip = "on") +
     scale_color_discrete(name = "Pre-PCR coefficient of variation") +
     theme(
       legend.position = "top",
       strip.background = element_rect(fill = "#f2f2f2")
     )
}
```

### Log-normally distributed pre-PCR noise
```{r}
if (rerun_computations) {
  nondetection_probs_lnorm <- compute_nondetection_probs(
    example_lambda_values = seq(0.001,15,by=0.1),
    example_c = 0.519e-6 * 30,
    example_total_partitions = 25000,
    example_cv = c(0, 0.5, 1, 1.5, 2),
    noise_dist = "lognormal",
    example_replicates = c(1,2),
    seed = 0
  )
  
  nondetection_probs_lnorm <- nondetection_probs_lnorm |> mutate(n_replicates_label = paste(ifelse(n_replicates==1,"(i)","(ii)"),n_replicates,ifelse(n_replicates==1,"replicate","replicates")))
  
  saveRDS(nondetection_probs_lnorm, here::here("data", "results", "nondetection_probs_lnorm.rds"))
}
```

```{r}
nondetection_probs_lnorm <- readRDS(here::here("data", "results", "nondetection_probs_lnorm.rds"))
```

```{r}
(ggplot(nondetection_probs_lnorm, aes(x = lambda)) +
  geom_line(aes(y = simulated, color = cv), alpha = 0.5) + 
  geom_line(aes(y = theoretical, color = cv))
) |> 
  format_nondetect_plot()

ggsave(here::here("figures", "dPCR paper", "prob_nondetection_lnorm.pdf"), width = 10, height = 5)
```

### Gamma distributed pre-PCR noise
```{r}
if (rerun_computations) {
  nondetection_probs_gamma <- compute_nondetection_probs(
    example_lambda_values = seq(0.001,15,by=0.1),
    example_c = 0.519e-6 * 30,
    example_total_partitions = 25000,
    example_cv = c(0, 0.5, 1, 1.5, 2),
    noise_dist = "gamma",
    example_replicates = c(1,2),
    seed = 0
  )
  
  nondetection_probs_gamma <- nondetection_probs_gamma |> mutate(n_replicates_label = paste(ifelse(n_replicates==1,"(i)","(ii)"),n_replicates,ifelse(n_replicates==1,"replicate","replicates")))
  
  saveRDS(nondetection_probs_gamma, here::here("data", "results", "nondetection_probs_gamma.rds"))
}
```

```{r}
nondetection_probs_gamma <- readRDS(here::here("data", "results", "nondetection_probs_gamma.rds"))
```

```{r}
prob_nondetection_plot_gamma <- (ggplot(nondetection_probs_gamma, aes(x = lambda)) +
  geom_line(aes(y = simulated, color = cv), alpha = 0.5) + 
  geom_line(aes(y = theoretical, color = cv))
) |> 
  format_nondetect_plot()

ggsave(prob_nondetection_plot_gamma, filename = here::here("figures", "dPCR paper", "prob_nondetection_gamma.pdf"), width = 10, height = 5)
```

### Example probabilities in main manuscript

Non-detection probability without pre-PCR noise
```{r}
est_nondetection_prob(5, c = 1.73e-5, n = 25000, n_replicates = 1)
```

Non-detection probability with pre-PCR noise
```{r}
est_nondetection_prob_pre(5, c = 1.73e-5, n = 25000, cv = 0.5, n_replicates = 1, noise_dist = "gamma")
```

## Combined figure
```{r, fig.width = 10, fig.height = 8}
mainplot <- cowplot::plot_grid(plotlist = cowplot::align_plots(cv_pre_plot_gamma + theme(legend.position = "none"), prob_nondetection_plot_gamma  + theme(legend.position = "none"), align = "v", axis = "l"), ncol = 1, labels = "AUTO")
plotlegend <- cowplot::get_plot_component(cv_pre_plot_gamma  + theme(legend.position = "top"), pattern = "guide-box-top")

cowplot::plot_grid(plotlegend, mainplot, ncol = 1, rel_heights = c(0.05, 0.95))

ggsave(here::here("figures", "dPCR paper", "cv_pre_non_detect_gamma.pdf"), width = 10, height = 8)
```