```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
here::i_am("notebooks/dPCR paper/noise_model_comparison.Rmd")
```

```{r}
library(targets)
library(dplyr)
library(EpiSewer)
library(ggplot2)
library(stringr)
source("code/local_config.R")
source("code/pipeline/utils_pipeline.R")
```

```{r}
setup_pipeline("_noise_model_comparison")
```

## Preparation

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Run pipeline

All results are already available and will be loaded. If you want to rerun the models, set `submit = TRUE`.
```{r, include = FALSE}
run_pipeline(submit = F, targets_proj_name = "_noise_model_comparison", run_cluster = FALSE)
if(length(tar_read(jobs_EpiSewer_invalid))>0) {
  warning("There are invalid jobs.")
}
```

```{r}
tar_load(job_EpiSewer_submission)

job_EpiSewer_submission |> 
  mutate(total = 1, .before = exists) |>
  summarise(across(total:submitted, sum, na.rm = T))
```
## Load results

```{r}
setup_pipeline("_noise_model_comparison")
tar_load(job_EpiSewer_result)
```

## Main figure
```{r}
intervals_plot <- c(0.5, 0.95)
conc_plot_f <- function(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values) {
  plot_concentration(
    res_list,
    measurements = PCR_data_plot[, c("date", "gc_per_mlww")],
    median = T, concentration_col = "gc_per_mlww",
    facet_models = TRUE, facet_direction = "cols",
    forecast = T, forecast_horizon = f_horizon,
    intervals = intervals_plot
    ) + 
    scale_color_manual(values = color_values) +
    scale_fill_manual(values = color_values) +
    ggpattern::scale_pattern_color_manual(values = color_values) +
    ggpattern::scale_pattern_fill_manual(values = color_values) +
    coord_cartesian(ylim = c(0, 510), xlim = c(as.Date("2022-10-20"), NA)) + 
    scale_y_continuous(expand = expansion(add = c(0.5,0))) +
    theme(
      axis.title.x = element_blank()
    ) +
    geom_point(
      data = tar_read(data_PCR_select, 1+5*(1 + (i %/% 6)))[[1]][date <= last_estim_date, c("date", "gc_per_mlww")],
      aes(x=date, y=gc_per_mlww),
      color = "black", shape = 4
      ) + 
    geom_point(
      data = tar_read(data_PCR_select, 1+5*(1 + (i %/% 6)))[[1]][date > last_estim_date & date <= last_estim_date + f_horizon, c("date", "gc_per_mlww")],
      aes(x=date, y=gc_per_mlww),
      color = "black", shape = 8
      ) + 
    ylab("Concentration [gc/mL]") +
    geom_vline(xintercept = last_estim_date, color = "grey", linetype = "dotted")
}

load_plot_f <- function(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values) {
  plot_load(
    res_list,
    median = T,
    facet_models = TRUE, facet_direction = "cols",
    forecast = T, forecast_horizon = f_horizon,
    intervals = intervals_plot
    ) + 
    scale_color_manual(values = color_values) +
    scale_fill_manual(values = color_values) +
    ggpattern::scale_pattern_color_manual(values = color_values) +
    ggpattern::scale_pattern_fill_manual(values = color_values) +
    coord_cartesian(ylim = c(0, 5e13), xlim = c(as.Date("2022-10-20"), NA)) +
    scale_y_continuous(expand = expansion(add = c(0.5,0))) +
    theme(
      axis.title.x = element_blank()
    ) +
    geom_vline(xintercept = last_estim_date, color = "grey", linetype = "dotted")
}

inf_plot_f <- function(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values){
  plot_infections(
    res_list,
    median = T,
    facet_models = TRUE, facet_direction = "cols",
    seeding = T,
    forecast = T, forecast_horizon = f_horizon,
    intervals = intervals_plot
    ) + 
    scale_color_manual(values = color_values) +
    scale_fill_manual(values = color_values) +
    ggpattern::scale_pattern_color_manual(values = color_values) +
    ggpattern::scale_pattern_fill_manual(values = color_values) +
    coord_cartesian(ylim = c(0, NA), xlim = c(as.Date("2022-10-20"), NA)) +
    scale_y_continuous(expand = expansion(add = c(0.5,0))) +
    theme(
      axis.title.x = element_blank()
    ) +
    geom_vline(xintercept = last_estim_date, color = "grey", linetype = "dotted")
}

R_plot_f <- function(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values){
  plot_R(
    res_list,
    median = T,
    facet_models = TRUE, facet_direction = "cols",
    seeding = T,
    forecast = T, forecast_horizon = f_horizon,
    intervals = intervals_plot
    ) + 
    scale_color_manual(values = color_values) +
    scale_fill_manual(values = color_values) +
    ggpattern::scale_pattern_color_manual(values = color_values) +
    ggpattern::scale_pattern_fill_manual(values = color_values) +
    coord_cartesian(ylim = c(0, NA), xlim = c(as.Date("2022-10-20"), NA)) +
    theme(
      axis.title.x = element_blank()
    ) +
    geom_vline(xintercept = last_estim_date, color = "grey", linetype = "dotted") +
    scale_y_continuous(name = expression(Reproduction~number~R[t]))
}
```

Main comparison of noise models
```{r fig.width = 15, fig.height = 12}
color_values <- c("#d31f11","#f47a00","#007191")
f_horizon <- 7

for (i in 1:11) {
  res_list <- job_EpiSewer_result[c(1:3)+i*4]
  names(res_list) <- c("1) Normal likelihood", "2) Log-Normal likelihood", "3) dPCR likelihood")
  
  PCR_data_plot <- tar_read(data_PCR_select, i+1)[[1]]
  last_estim_date <- max(PCR_data_plot$date, na.rm = T)

  conc_plot <- conc_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  load_plot <- load_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  inf_plot <- inf_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  R_plot <- R_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  
  full_plot <- cowplot::plot_grid(
    plotlist = cowplot::align_plots(conc_plot, load_plot, R_plot, align = 'v', axis = 'l'), ncol = 1, labels = "AUTO"
    )
  print(full_plot)
  
  file_id <- stringr::str_replace_all(paste(
    res_list[[1]]$job$selection$wwtp, res_list[[1]]$job$selection$target, res_list[[1]]$job$selection$date_select[[1]]["to"], collapse = "_"
    ), "[^[:alnum:]]", "_")
  ggsave(full_plot, filename = here::here("figures", "dPCR paper",paste0("noise_model_comparison_", file_id, ".pdf")), width = 10, height = 8)
}
```

Comparison of informed vs uninformed dPCR model
```{r fig.width = 15, fig.height = 12}
color_values <- c("#007191", "#8f008f")
f_horizon <- 7

for (i in 3:5) {
  res_list <- job_EpiSewer_result[c(3,4)+i*4]
  names(res_list) <- c("1) dPCR parameters unknown", "2) dPCR parameters known")
  
  PCR_data_plot <- tar_read(data_PCR_select, i+1)[[1]]
  last_estim_date <- max(PCR_data_plot$date, na.rm = T)

  conc_plot <- conc_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  load_plot <- load_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  inf_plot <- inf_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  R_plot <- R_plot_f(res_list, PCR_data_plot, f_horizon, last_estim_date, color_values)
  
  full_plot <- cowplot::plot_grid(
    plotlist = cowplot::align_plots(conc_plot, load_plot, R_plot, align = 'v', axis = 'l'), ncol = 1, labels = "AUTO"
    )
  print(full_plot)
  
  file_id <- stringr::str_replace_all(paste(
    res_list[[1]]$job$selection$wwtp, res_list[[1]]$job$selection$target, res_list[[1]]$job$selection$date_select[[1]]["to"], collapse = "_"
    ), "[^[:alnum:]]", "_")
  ggsave(full_plot, filename = here::here("figures", "dPCR paper",paste0("dPCR_parameters_known_unknown", file_id, ".pdf")), width = 10, height = 8)
}
```

## Evaluation

### Forecast score
```{r}
i <- 3
f_horizon <- 7
lapply(c(1:3)+i*4, function(sub_i) {
forecast_dates <- seq(job_EpiSewer_result[[sub_i]]$job$metainfo$T_end_date + 1, length.out = f_horizon, by = "1 day")

true_conc <- tar_read(data_PCR_select)[[ifelse(3<=6,6,12)]][
  date %in% forecast_dates
  ]

observed_dates <- forecast_dates[forecast_dates %in% true_conc$date]
observed_dates <- true_conc$date[true_conc$date %in% observed_dates]
# get index of each observed date in forecast_dates
observed_date_indices <- sapply(observed_dates, function(x) which(forecast_dates == x))

forecast_conc <- job_EpiSewer_result[[sub_i]]$fitted$draws("predicted_concentration_forecast")
forecast_conc <- t(posterior::as_draws_matrix(forecast_conc))

for (t in 1:nrow(forecast_conc)) {
  forecast_conc[t, is.na(forecast_conc[t,])] <- sample(forecast_conc[t, !is.na(forecast_conc[t,])], size = sum(is.na(forecast_conc[t,])), replace = T)
}
#print(as.vector(rowMeans(forecast_conc, na.rm = T)))

mean(scoringutils::crps_sample(
  true_conc$gc_per_mlww,
  forecast_conc[observed_date_indices,]
))
})
```

### Interval coverage
```{r}
i <- 3

res_list <- job_EpiSewer_result[c(1:3)+i*4]
names(res_list) <- c("1) Normal likelihood", "2) Log-Normal likelihood", "3) dPCR likelihood")

conc_pred <- bind_rows(lapply(1:3, function(sub_i) {
res_list[[sub_i]]$summary$concentration[type == "estimate", ] |> 
  mutate(model = names(res_list)[sub_i], .before = 1)
}))

conc_true <- tar_read(data_PCR_select)[[ifelse(3<=6,6,12)]][date %in% conc_pred$date,]

# join conc true and conc pred
conc_score <- merge(conc_pred, conc_true, by = "date") |> 
  mutate(
    coverage_0.5 = ifelse(gc_per_mlww >= lower_0.5 & gc_per_mlww <= upper_0.5, 1, 0),
    coverage_0.8 = ifelse(gc_per_mlww >= lower_0.8 & gc_per_mlww <= upper_0.8, 1, 0),
    coverage_0.95 = ifelse(gc_per_mlww >= lower_0.95 & gc_per_mlww <= upper_0.95, 1, 0)
    ) |> group_by(model)

# all
conc_score |> 
  summarise(across(starts_with("coverage_"), function(x) mean(x, na.rm = T)))

# low concentrations
conc_score |> 
  filter(date < as.Date("2022-12-01")) |> 
  summarise(across(starts_with("coverage_"), function(x) mean(x, na.rm = T)))

# high concentrations
conc_score |> 
  #filter(mean > 100) |> 
  filter(date >= as.Date("2022-12-01")) |> 
  summarise(across(starts_with("coverage_"), function(x) mean(x, na.rm = T)))

conc_pred |> 
  group_by(model) |>
  filter(date >= as.Date("2022-12-01")) |>
  mutate(
    next_consistent = median > lag(lower_0.95) & median < lag(upper_0.95)
  ) |> 
  summarize(share_next_consistent = sum(next_consistent, na.rm = T)/sum(!is.na(next_consistent)))
```

## Estimated parameters

Prior-posterior plot

```{r fig.width = 10, fig.height = 5}
i <- 3

plot_prePCR_cv <- plot_prior_posterior(job_EpiSewer_result[[i]], "nu_upsilon_a") + coord_cartesian(xlim = c(0,2)) + 
  theme(
  #axis.title.y = element_blank()
) + 
  ylab("") +
  xlab(expression(nu[pre])) +
  scale_y_continuous(expand = expansion(mult = c(-0.01,0.05)), labels = scales::label_scientific())

plot_total_partitions <- plot_prior_posterior(job_EpiSewer_result[[i]], "nu_upsilon_b_mu") + geom_vline(xintercept = 25468, color = "red") +
  xlab(expression(mu[m]))  +
  scale_y_continuous(expand = expansion(mult = c(-0.01,0.05)),labels = scales::label_scientific())

plot_conversion_factor <- plot_prior_posterior(job_EpiSewer_result[[i]], "nu_upsilon_c") + geom_vline(xintercept = 1.73e-5, color = "red") + 
    theme(
  axis.title.y = element_blank()
) +
  xlab(expression(c)) +
  scale_y_continuous(expand = expansion(mult = c(-0.01,0.05)), labels = scales::label_scientific())

cowplot::plot_grid(plot_total_partitions, plot_conversion_factor, plot_prePCR_cv, nrow = 1, labels = "AUTO")

ggsave(filename = here::here("figures", "dPCR paper", "noise_model_parameters.pdf"), width = 10, height = 3)
```

Interpretation of conversion factor
```{r}
i <- 3
nu_upsilon_a <- job_EpiSewer_result[[i]]$fitted$summary("nu_upsilon_a")

source(here::here("code", "utils_dPCR_statistics.R"))
source(here::here("code", "utils_dists.R"))

est_cv_pre_lnorm(lambda = 100, n = 25468, c = 1.73e-5, cv = 0.172, n_replicates = 1)

est_cv_pre_taylor(lambda = 100, n = 25468, c = 1.73e-5, cv = 0, n_replicates = 1)
```
## Pruning

```{r}
prune_results(remove_all_outputs = TRUE, remove_all_jobfiles = TRUE)
```

